Viscous Transport in Eroding Porous Media

### 2019-10-01 docs/transport/erosion_transport.tex
NM

Flow in porous media is further complicated when boundaries dynamically
evolve in response to the fluid flow. This type of coupling between the
geometry and the flow occurs, for example, in applications involving
melting~\cite{bec-vis1998, rycroft2016asymmetric, jambon2018singular,
favier2019rayleigh, morrow2019moving},
dissolution~\cite{kan-zha-che-he2002, mac2015shape, moo2017,
wykes2018self}, deposition~\cite{joh-eli1995, hewett2018modelling},
biofilm growth~\cite{tan-val-wer2015}, and crack
formation~\cite{cho2019crack}. We focus on erosion, a fluid-mechanical
process that is prevalent in many geophysical, hydrological, and
industrial applications~\cite{berhanu2012shape, hewett2017evolution,
lachaussee2018competitive, lopez2018cfd, allen2019sde, amin2019role}.
When a porous geometry erodes, certain qualitative characteristics are
unveiled that affect transport through the geometry.  For example, an
eroded geometry may contain channels of high porosity, which, though few
in number and modest in volume fraction, transmit a large portion of the
flux. This arrangement results in velocities that vary over several
orders of magnitude~\cite{all-hea-lab-rei2002}.  Moreover,
channelization creates heterogeneous and anisotropic medium properties,
which affect the transport of tracers such as
contaminants~\cite{cve-che-wen1996, dag1987, kon-bre1978} and
heat~\cite{nil-sto1990, ree-sto1995}.

This study consists of two main undertakings: first, high-fidelity
simulations of eroding porous media, and, second, characterization of
tracer transport through the resulting eroded geometries. Our modeling
efforts build on previous work~\cite{ris-moo-chi-she-zha2012,
moore2013self, moo2017}, in particular recent numerical methods
developed to simulate erosion in the regime of Stokes flow
regime~\cite{qua-moo2018}. We, however, make key improvements to the
numerical methods from~\cite{qua-moo2018} to enable simulations of more
realistic, dense suspensions of eroding grains
(Figures~\ref{fig:Eroding50vel} and~\ref{fig:Eroding50vort}).  To
characterize transport through these configurations, we examine
coarse-grained variables through statistical analysis of tracer
trajectories.

@@

Owing to the scales present in groundwater flow~\cite{bea1972}, we model
the hydrodynamics with the two-dimensional incompressible Stokes
equations.  Meanwhile, individual grains erode at a rate proportional to
the hydrodynamic shear stress~\cite{wan-fel2004,
ris-moo-chi-she-zha2012, moore2013self, par-izu2000}.  Since the
vorticity and the shear stress are equivalent on the grains in the zero
Reynolds number limit, we also compute the vorticity in the fluid bulk
for visualization purposes (Figure~\ref{fig:Eroding50vort}).  A variety
of numerical methods for the fluid solver are available including
lattice-gas or lattice Boltzmann~\cite{kop-kat-tim1996, dar-mcc1998},
finite volumes~\cite{suo-liu-gan2019, den-ica-hid2018,
sie-ili-pri-riv-gua2019}, finite differences~\cite{leb-ded-dav-bou2007,
knu-car2005}, and cellular automata~\cite{rot1988}.  However, simulating
erosion requires a method that is high-order, can resolve
nearly-touching arbitrarily shaped grains, and can accurately compute
derivatives of the velocity field.  Since the governing equations are
linear and homogeneous, we address these challenges by converting the
fluid equations to a boundary integral equation (BIE).

To compute stable simulations of erosion, we use high-order methods in
both space and time. The time integration method is unchanged from our
previous work~\cite{qua-moo2018}.  We apply a mild regularization and a
smoothing term to eliminate numerical instabilities caused by changes in
sign of the shear stress, and we use a stable second-order Runge-Kutta
method applied to the {\thL} coordinates~\cite{hou-low-she1994} of the
eroding grains.  In this work, we introduce a new quadrature method to
resolve dense suspensions of eroding grains.  The accuracy of the
trapezoid rule, which we used in our previous work, is adequate for
bodies that are sufficiently separated~\cite{tre-wei2014}.  However, it
loses accuracy whenever grains are in close contact.  One of the
earliest works to address quadrature methods for nearly-singular
integrands was developed by Baker and Shelley~\cite{bak-she1986}, and in
recent years, many other near-singular integration schemes have
followed~\cite{oja-tor2015, kli-tor2018, hel-oja2008a, bea-yin-wil2016,
bea-lai2001, klo-bar-gre-one2013}.  We use a Barycentric quadrature
method~\cite{bar2014, bar-wu-vee2015} since it is a non-intrusive
modification of the trapezoid rule, and the error is guaranteed to be
uniformly bounded.  We extend the original quadrature method to compute
the velocity gradient, which is needed to evaluate shear stress and
fluid vorticity.

@@

To characterize the transport, particle trajectories must be computed.
Depending on the application, microscale transport can be modelled as
pure advection~\cite{dea-qua-bir-jua2018, leb-ded-dav-bou2007,
cve-che-wen1996, puy-gou-den2019},
advection-diffusion~\cite{cus-hu-den1995, dag1987, den-ica-hid2018}, or
with a random walk~\cite{saf1959, bij-blu2006, ber-sch-sil2000}.  In
this work, we assume trajectories to be governed by pure advection
(ie.~no diffusion), so the particle trajectories are identical to the
streamlines. We compute trajectories $\ss(t)$ that are initialized at
$\ss_0$ by solving the advection equation
\begin{align}
  \frac{d\ss}{dt} = \uu(\ss,t), \quad \ss(0) = \ss_0,
  \label{eqn:tracers}
\end{align}
where $\uu$ is the fluid velocity.  Since there is no stiff diffusive
term in~\eqref{eqn:tracers}, we use a fourth-order explicit Runge-Kutta
time stepping method.  In addition, the Barycentric quadrature rule is
used to accurately compute trajectories that are close to an eroding
grain. We note that an alternative approach to compute trajectories is
to form a distribution of velocities $\uu$ and use a continuous time
random walk~\cite{mon-wei1965, den-cor-sch-ber2004, leb-den-car2008,
ber-cor-den-sch2006}.

@@

Our second contribution is computing and analyzing trajectories in
eroded geometries.  These trajectories are formed by combining the
Barycentric quadrature rule and a fourth-order Runge-Kutta method. Then,
they are used to compute tortuosities and dispersion rates of eroded
geometries. Finally, we analyze the effect of erosion on the pore size
distribution.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Governing Equations}
\label{sec:formulation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We start by defining the main variables used to model erosion.  We only
briefly summarize the model, and a more detailed description can be
found in previous work~\cite{qua-moo2018}.  We consider flows inside a
confined geometry $\Omega$ that contains $M$ eroding bodies with
boundaries $\gamma_\ell$, $\ell = 1,\ldots,M$.  The boundary of the
fluid domain is $\bd \Omega = \Gamma \cup \gamma_1 \cup \cdots \cup
\gamma_M$, where $\Gamma$ is the outer boundary, taken to be a slightly
smoothed version of $[-3,3] \times [-1,1]$.  We only place eroding
bodies in $[-1,1] \times [-1,1]$ so as to create a buffer region for the
flow profile imposed at the inlet to transition to the more complex flow
intervening between the bodies. On each eroding body, $\nn$ is the
normal vector pointing into the body, and $\ss$ is the unit tangent
vector pointing in the counterclockwise direction.  The normal vector on
$\Gamma$ is the outward unit normal.  Neglecting inertial forces, the
governing equations are
\begin{equation}
\label{eqn:erosionModel}
  \begin{split}
    \mu \Delta \uu = \grad p, &\hspace{20pt} \xx \in \Omega, \gap 
      &&\mbox{\em conservation of momentum}, \\
    \grad \cdot \uu = 0, &\hspace{20pt} \xx \in \Omega, \gap 
      &&\mbox{\em conservation of mass}, \\
    \uu = \mathbf{0}, &\hspace{20pt} \xx \in \gamma, \gap 
      &&\mbox{\em no slip on the eroding bodies}, \\
    \uu = \UU, &\hspace{20pt} \xx \in \Gamma, \gap 
      &&\mbox{\em outer wall velocity}, \\
    \Vn = \CE \, \abs{\tau}, &\hspace{20pt} \xx \in \gamma,
      &&\mbox{\em erosion model}.
  \end{split}
\end{equation}
Here, $\uu$ is the fluid velocity, $p$ is the pressure, $\UU$ is a
prescribed Hagen-Poiseuille velocity field, $\Vn$ is the normal velocity
of $\gamma$, and
\begin{align}
  \tau = -(\nabla \uu + \nabla \uu^T) \nn \cdot \ss
  \label{eqn:shearStress}
\end{align}
is the shear stress on $\gamma$. We simulate erosion by alternating
between solving the fluid equations nd advancing the grains according to
the erosion rate for a small time step. The strength of $\UU$ is
adjusted at each time step to achieve a constant pressure drop across
the channel.

>>>>

Flow in porous media is further complicated when boundaries dynamically
evolve in response to the fluid flow. This type of coupling between the
geometry and the flow occurs, for example, in applications involving
melting~\cite{bec-vis1998, rycroft2016asymmetric, jambon2018singular,
favier2019rayleigh, morrow2019moving},
dissolution~\cite{kan-zha-che-he2002, mac2015shape, moo2017,
wykes2018self}, deposition~\cite{joh-eli1995, hewett2018modelling},
biofilm growth~\cite{tan-val-wer2015}, and crack
formation~\cite{cho2019crack}. We focus on erosion, a fluid-mechanical
process that is prevalent in many geophysical, hydrological, and
industrial applications~\cite{berhanu2012shape, hewett2017evolution,
lachaussee2018competitive, lopez2018cfd, allen2019sde, amin2019role}.
When a porous geometry erodes, certain qualitative characteristics are
unveiled that affect transport through the geometry.  For example, an
eroded geometry may contain channels of high porosity, which, though few
in number and modest in volume fraction, transmit a large portion of the
flux \cite{qua-moo2018}. This arrangement results in velocities that vary over several
orders of magnitude~\cite{all-hea-lab-rei2002}.  Moreover,
channelization creates heterogeneous and anisotropic medium properties,
which affect the transport of tracers such as
contaminants~\cite{cve-che-wen1996, dag1987, kon-bre1978} and
heat~\cite{nil-sto1990, ree-sto1995}.

This study consists of two main undertakings: first, high-fidelity
simulations of eroding porous media, and, second, characterization of
tracer transport through the resulting eroded geometries. Our modeling
efforts build on previous work~\cite{ris-moo-chi-she-zha2012,
moore2013self, moo2017}, in particular recent numerical methods
developed to simulate erosion in the Stokes-flow
regime~\cite{qua-moo2018}. We, however, make key improvements to the
numerical methods from~\cite{qua-moo2018} to enable simulations of more
realistic, dense suspensions of eroding grains
(Figures~\ref{fig:Eroding50vel} and~\ref{fig:Eroding50vort}).  To
characterize transport through these configurations, we examine
coarse-grained variables through statistical analysis of tracer
trajectories.

@@

Owing to the scales present in groundwater flow~\cite{bea1972}, we model
the hydrodynamics with the two-dimensional incompressible Stokes
equations.  Meanwhile, individual grains erode at a rate proportional to
the hydrodynamic shear stress~\cite{wan-fel2004,
ris-moo-chi-she-zha2012, moore2013self, par-izu2000}.  Since the
vorticity and the shear stress are equivalent on \nick{solid boundaries}, we also compute the fluid vorticity for visualization purposes (Figure~\ref{fig:Eroding50vort}).  A variety of numerical methods for the fluid solver are available including
lattice-gas or lattice Boltzmann~\cite{kop-kat-tim1996, dar-mcc1998},
finite volumes~\cite{suo-liu-gan2019, den-ica-hid2018,
sie-ili-pri-riv-gua2019}, finite differences~\cite{leb-ded-dav-bou2007,
knu-car2005}, and cellular automata~\cite{rot1988}.  However, simulating
erosion requires a method that is high-order, can resolve
nearly-touching arbitrarily shaped grains, and can accurately compute
derivatives of the velocity field.  Since the governing equations are
linear and homogeneous, we address these challenges by converting the
fluid equations to a boundary integral equation (BIE).

To compute stable simulations of erosion, we use high-order methods in
both space and time. The time integration method is unchanged from
previous work~\cite{qua-moo2018}.  We apply a mild regularization and a
smoothing term to eliminate numerical instabilities that can be triggered by changes in
sign of the shear stress, and we use a stable second-order Runge-Kutta
method applied to the {\thL} coordinates~\cite{hou-low-she1994} of the
eroding grains.  In this work, we introduce a new quadrature method to
resolve dense suspensions.  The accuracy of the
trapezoid rule, which was used in previous work, is adequate for
bodies that are sufficiently separated~\cite{tre-wei2014}.  However, it
loses accuracy whenever grains are in close contact.  One of the
earliest works to address quadrature methods for nearly-singular
integrands was developed by Baker and Shelley~\cite{bak-she1986}, and in
recent years, many other near-singular integration schemes have
followed~\cite{oja-tor2015, kli-tor2018, hel-oja2008a, bea-yin-wil2016,
bea-lai2001, klo-bar-gre-one2013}.  We use a Barycentric quadrature
method~\cite{bar2014, bar-wu-vee2015} since it is a non-intrusive
modification of the trapezoid rule, and the error is guaranteed to be
uniformly bounded.  We extend the original quadrature method to compute
the velocity gradient, which is needed to evaluate shear stress and
fluid vorticity.

@@

To characterize transport \nick{through the resulting configurations}, particle trajectories must be computed. Depending on the application, microscale transport can be modelled as
pure advection~\cite{dea-qua-bir-jua2018, leb-ded-dav-bou2007,
cve-che-wen1996, puy-gou-den2019},
advection-diffusion~\cite{cus-hu-den1995, dag1987, den-ica-hid2018}, or
with a random walk~\cite{saf1959, bij-blu2006, ber-sch-sil2000}.  In
this work, we assume trajectories to be governed by pure advection
(ie.~no diffusion), so the particle trajectories are identical to the
streamlines. We compute trajectories $\ss(t)$ that are initialized at
$\ss_0$ by solving the advection equation
\begin{align}
  \frac{d\ss}{dt} = \uu(\ss,t), \quad \ss(0) = \ss_0,
  \label{eqn:tracers}
\end{align}
where $\uu$ is the fluid velocity.  Since there is no stiff diffusive
term in~\eqref{eqn:tracers}, we use a fourth-order explicit Runge-Kutta
time stepping method.  In addition, the Barycentric quadrature rule is
used to accurately compute trajectories that are close to an eroding
grain. We note that an alternative approach to compute trajectories is
to form a distribution of velocities $\uu$ and use a continuous time
random walk~\cite{mon-wei1965, den-cor-sch-ber2004, leb-den-car2008,
ber-cor-den-sch2006}.

@@

Our second contribution is computing and analyzing trajectories in
eroded geometries.  These trajectories are formed by combining the
Barycentric quadrature rule and a fourth-order Runge-Kutta method. Then,
they are used to compute tortuosities and dispersion rates of eroded
geometries. Finally, we examine the effect of erosion on the pore size
distribution.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Governing Equations}
\label{sec:formulation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We start by defining the main variables used to model erosion.  We only
briefly summarize the model, and a more detailed description can be
found in previous work~\cite{qua-moo2018}.  We consider flows inside a
confined geometry $\Omega$ that contains $M$ eroding bodies with
boundaries $\gamma_\ell$, $\ell = 1,\ldots,M$.  The boundary of the
fluid domain is $\bd \Omega = \Gamma \cup \gamma_1 \cup \cdots \cup
\gamma_M$, where $\Gamma$ is the outer boundary, taken to be a slightly
smoothed version of $[-3,3] \times [-1,1]$.  We only place eroding
bodies in $[-1,1] \times [-1,1]$ so as to create a buffer region for the
flow profile imposed at the inlet to transition to the more complex flow
intervening between the bodies. On each eroding body, $\nn$ is the
normal vector pointing into the body, and $\ss$ is the unit tangent
vector pointing in the counterclockwise direction.  The normal vector on
$\Gamma$ is the outward unit normal.  Neglecting inertial forces, the
governing equations are
\begin{equation}
\label{eqn:erosionModel}
  \begin{split}
    \mu \Delta \uu = \grad p, &\hspace{20pt} \xx \in \Omega, \gap 
      &&\mbox{\em conservation of momentum}, \\
    \grad \cdot \uu = 0, &\hspace{20pt} \xx \in \Omega, \gap 
      &&\mbox{\em conservation of mass}, \\
    \uu = \mathbf{0}, &\hspace{20pt} \xx \in \gamma, \gap 
      &&\mbox{\em no slip on the eroding bodies}, \\
    \uu = \UU, &\hspace{20pt} \xx \in \Gamma, \gap 
      &&\mbox{\em outer wall velocity}, \\
    \Vn = \CE \, \abs{\tau}, &\hspace{20pt} \xx \in \gamma,
      &&\mbox{\em erosion model}.
  \end{split}
\end{equation}
Here, $\uu$ is the fluid velocity, $p$ is the pressure, $\UU$ is a
prescribed Hagen-Poiseuille velocity field, $\Vn$ is the normal velocity
of $\gamma$, and
\begin{align}
  \tau = -(\nabla \uu + \nabla \uu^T) \nn \cdot \ss
  \label{eqn:shearStress}
\end{align}
is the shear stress on $\gamma$. We simulate erosion by alternating
between solving the fluid equations \nick{and} advancing the grains according to
the erosion rate for a small time step. The strength of $\UU$ is
adjusted at each time step to achieve a constant pressure drop across
the channel.
